A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-07-16 paper/paper.tex
Minor change

\usepackage{amsmath, amssymb, mathrsfs, url, bookmark, tikz, tikz-cd, mathpazo, enumitem, xargs, multirow, float, tabularx, environ, booktabs, setspace, makecell, colortbl, pdflscape, flushend, minted, art.cls/colorpal, art.cls/ct, art.cls/sset, art.cls/lim, art.cls/joinargs}

@@

\subsection*{Fibered vs indexed presentation of semi-simplicial and semi-cubical sets}
A family of sets can commonly be represented in two ways: as a family properly speaking, indexed over a set $S$, or as a set $T$ together with a map specifying for each element of the set $T$ on which index it depends in $S$. In the first case, we call it an \emph{indexed} presentation. In the latter case, the set associated to a given index is the fiber, i.e. inverse image, of the index under the map, so we call it a \emph{fibered} presentation. The two presentations are equivalent and the equivalence can be phrased concisely in the language of homotopy type theory~\cite{hottbook} as a fibered/indexed correspondence\footnote{At a discussion on the Zulip channel for the HoTT community, alternative names fibration/family correspondence and unbundled/bundled correspondence were proposed, among others. In our experience, the opposition fibered/indexed is spontaneously understood by many listeners. In particular, it echoes to the Grothendieck construction of fibered categories from indexed categories. The most elementary instance of the correspondence, with $\Type$ instead of $\U$, that is $(\Sigma S: \Type. (S \rightarrow T)) \simeq (T \rightarrow \Type)$, is sometimes called ``Grothendieck construction for dummies'' (its proof requires univalence, see~\cite{hottbook}).}
\begin{equation*}
  \mbox{(fibered)}\qquad(\Sigma S: \U. (S \rightarrow T)) ~\simeq~ (T \rightarrow \U)\qquad \mbox{(indexed)}
\end{equation*}
where $\U$ represents sets in homotopy type theory as the subset of types with a unique proof of equality (within some universe).

@@

In category-theoretic terms, the indexi set of the family $X_n$ in the indexed presentation is called a \emph{matching object}. The indexed presentation of presheaves over a direct category can then be identified with a subclass of ordinary presheaves known as Reedy fibrant where Reedy fibrancy precisely expresses that the set $X_n$ in the fibered presentation is a $\Sigma$-type over an instance up to isomorphism of the matching object at level $n$~\citep{shulman14,kraus17}.

@@

%% Notations
% simplicial, augmented, where
% Γ           Γ+         Augmented Simplicial Combinatorics through Category Theory
%                        José Manuel García-Calcines, María Teresa Rivas-Rodríguez, Luis Javier Hernández-Paricio, 2022

% semi-augmented, where
% ∆_s        HOMOLOGY IN RELATIVE SEMI-ABELIAN CATEGORIES, JULIA GOEDECKE, 2011

% simplicial, semi-simplicial, where
% Δ           Δ_inj            SEMI-SIMPLICIAL SPACES, JOHANNES EBERT AND OSCAR RANDAL-WILLIAMS, 2018
% Δ           Δ_inj            PhD, Josep Esquirol Esteve, 2019
% Δ           Δ'               Strict n-categories and augmented directed complexes model homotopy types, Andrea Gagna, 2018
% Δ           Δ_inj, Δ_{inj,aug} Ms Th. Jonas Dallend ̈orfer, 2019

@@

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj(\DeltaPlus) & := \Nat                                                                     \\
      \Hom(p, n)       & := \{l \in [0, \kstar]^n \mid \text{number of $\kstar$ in $l = p$}\}        \\
      g \circ f        & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                        \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id              & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom(n, n)$}
    \end{align*}
  \end{figure}
\end{definition}

@@

\begin{definition}[$\DeltaPlus^n$]
  The standard augmented $n$-semi-simplex $\DeltaPlus^n$ is defined as what is called the Yoneda embedding of $n \in \Obj(\DeltaPlus)$:
  \begin{align*}
    \DeltaPlus^n    & : \Set_{\DeltaPlus}        \\
    \DeltaPlus^n(p) & := \Hom(p, n)              \\
    \DeltaPlus^n(f) & := \lambda g .\, g \circ f
  \end{align*}
\end{definition}

@@

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj(\Cube) & := \Nat                                                                                 \\
      \Hom(p, n)  & := \{l \in [L, R, \kstar]^n \mid \text{number of $\kstar$ in $l = p$}\}                 \\
      g \circ f   & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                               \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                           \\
        a :: (g' \circ f') & \text{if $g = (\kstar :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id         & := [\kstar, \ldots, \kstar] \text{ $n$ times}
    \end{align*}
  \end{figure}

@@

\begin{example}[$\Cube^1$]
  $\Cube^1$ consists of two points, given by $\Hom(0, 1)$ and a line, given by $\Hom(1, 1)$.
  \begin{equation*}
    \begin{tikzcd}
      L \arrow[r, "\kstar", dash] & R
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\subsection*{\texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the generalization of augmented semi-simplicial sets and semi-cubical sets. To obtain this, we extend $\DeltaPlus$ and $\Cube$ in a straightforward manner into a category which we call $\nu$-semi-shape category.

\begin{definition}[$\nu$-semi-shape categroy]
  The definition of $\nu$-semi-shape category is shown below. Note that, if $g \circ f$ is well-defined, then the length of $f$ is less than that of $g$. It can be shown that composition is associative and that $\id$ is neutral.

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj       & := \Nat                                                                            \\
      \Hom(p, n) & := \{l \in (\nu \cup \kstar) \mid \text{number of $\kstar$ in $l = p$}\}           \\
      g \circ f  & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                         \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $R \in \nu$}                        \\
        a :: (g' \circ f') & \text{if $g = (\kstar :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases} \\
      \id        & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom(n, n)$}
    \end{align*}
  \end{figure}
\end{definition}

The morphisms of the $\nu$-semi-shape category are sequences of elements of a set $\nu$ of arbitrary cardinal, extended with $\kstar$:

@@

A $\nu$-set is thus a contravariant functor $\phi$ from the $\nu$-semi-shape category to $\Set$ and we call $n$-$\nu$-semi-shape an element of $\phi(n)$. As in the augmented semi-simplicial and semi-cubical cases, the standard $(n + 1)$-$\nu$-semi-shape is obtained by connecting together $\nu$ copies of the standard $n$-$\nu$-semi-shape with an extra copy stretched in the new dimension. We clarify in the next sections how this process of construction is similar to the parametricity translation developed for functional programming~\cite{reynolds83} and more generally for type theory~\cite{bernardy10, bernardy11, atkey14, bernardy15}.

\section{Preliminaries}
Martin-Löf's Type theory~\citep{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda~(\citeyear{agda23}), Coq~(\citeyear{coq22}) or Lean~(\citeyear{lean15}). In type theory, propositions are types and proofs are programs. A particularity of type theory is also that types and programs, hence propositions and proofs also, are considered modulo an equational theory called definitional equality.

Type theory is a flexible formalism supporting different models. Some models are based on topological spaces, where equality is interpreted as path, and substitutivity of equality as transport~\cite{kapulkin21}. These models support the univalence principle stating that equality of types mimics equivalence of types, leading to the development of Homotopy Type Theory~\cite{hottbook}. In type theory, types are organized in a hierarchy of universes written $\Type_m$ for $m$ a natural number. Main types in type theory are the types of dependent pairs, written $\Sigma a : A.\,B(a)$, the types of dependent functions, written $\Pi a:A.\,B(a)$, for $A$ a type and $B(a)$ a type dependent on the inhabitant $a$ of $A$, and the type of propositional equality, written $t = u$. We assume our type theory to also include a distinguished singleton type, written $\unittype$, and with inhabitant $\unitpoint$, the type of boolean values, called $\textsf{bool}$, and the type of natural numbers. We also write $\hd$ and $\tl$ the projections of dependent pairs, and $\refl$ for reflexivity. Logical propositions being types themselves, we use $\Pi$ to represent universal quantification and $\Sigma$ to represent existential quantification.

A notion of sets can be recovered in each universe as $\U[m]$, denoting the subtype of $\Type_m$ for which paths are degenerated, what can be expressed by the property of Uniqueness of Identity Proofs (UIP). Technically, this is expressed as a structure equipping a domain $\Dom$ with the property $\UIP$:
\begin{align*}
  \Dom & : \Type_m                                  \\
  \UIP & : \Pi x y: \Dom.\, \Pi p q: x = y.\, p = q
\end{align*}

@@

\section*{Comparison of fibered and indexed presentations}
%     - standard augmented triangle:
%       Inductive points := A | B | C.
%       Inductive lines := AB | BC | AC.
%       Inductive surface := ABC.

%       {X n := match n with 0 => unit | 1 => points | 2 => lines | 4 => surface | n => empty_set;
%        d {n} i (Hi:i<n) (x:X n) := match n, i, x with
%        | 0, _, _ => assert false
%        | 1, _, _ => tt
%        | 2, 0, AB => B
%        | 2, 1, AB => A
%        | 2, 0, BC => C
%        | 2, 1, BC => B
%        | 2, 0, AC => C
%        | 2, 1, AC => A
%        | 3, 0, ABC => BC
%        | 3, 1, ABC => AC
%        | 3, 2, ABC => AB
%        | n.+4, i, _ => assert false
%        end;
%        alpha n i j (Hi:i<n) (Hj:j<i) (x : X n) : d i Hi (d j _ x) = d (j-1) _ (d i _ x)
%        := match n, i, j, x with
%        | 0, _, _, _ => assert false
%        | 1, _, _, _ => assert false
%        | 2, _, _, _ => proof_irrelevance_of_unit : tt = tt
%        | 3, 1, 0, ABC => proof_irrelevance_in_points : C = C
%        | 3, 2, 0, ABC => proof_irrelevance_in_points : B = B
%        | 3, 2, 1, ABC => proof_irrelevance_in_points : A = A
%        | n.+4, i, _ => assert false
%        end
%        }
%        {hd tt := unit;
%         tl := {hd _ (* (tt;tt) *) := points; (* 3 points *)
%                tl := {hd _ (* ((tt;(tt;a));b) *) := unit; (* unique line over any a and any b *)
%                       tl := {hd _ (* ((tt;(tt;(a;(b;tt))));((c;tt);tt)) *) := unit; (* unique triangle over ab, bc, and ac *)
%                              tl := cofix s := {hd _ := empty_set; tl := s}}}}}
%        {hd _ := unit; tl :=
%        {hd _ := points; tl :=
%        {hd _ := unit; tl :=
%        {hd _ := unit; tl := cofix s := {hd _ := empty_set; tl := s}}}}} : AugmentedSemiSimplicial

%        standard n : AugmentedSemiSimplicial
%        := {hd _ := unit; tl :=
%           {hd _ := [0;n]; tl :=
%            unit unit unit ... n times ... unit; empty_set; ... }}
There are different ways to represent a family of sets, commonly known as the fibered and indexed representations. In type theory, they are equivalent, and this equivalence can be formulated as:
\begin{equation*}
  (\Sigma S: \U. (S \rightarrow T)) \simeq (T \rightarrow \U)
\end{equation*}

Let us apply this correspondence to the case of semi-cubical sets. A semi-cubical set can be represented as a sequence of fibrations, together with appropriate coherence conditions.
\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\partial^L" description, shift left=2] \arrow[l, "\partial^R" description, shift right=2] & X_2: \U \arrow[l, "\partial^{L\kstar}" description, shift left=6] \arrow[l, "\partial^{R\kstar}" description, shift left=2] \arrow[l, "\partial^{\kstar L}" description, shift right=2] \arrow[l, "\partial^{\kstar R}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

If we think of $X_1$ fibered over two copies of $X_0$ in an indexed way, we get $X_1$ of type $X_0 \rightarrow X_0 \rightarrow \U$. Continuing the process, we can think of $X_2$ fibered over four copies of $X_1$ themselves fibered over two copies of $X_0$ as a dependent type. We have a priori a dependency over eight copies of $X_0$ but these are actually four due to the coherence conditions on the functor (e.g. that $\partial^R \circ \partial^{R\kstar} = \partial^{(R\kstar)\circ(R)} = \partial^{RR} = \partial^{(\kstar R)\circ(R)} = \partial^R \circ \partial^{\kstar R}$, etc.). Based on these coherence conditions, the types of $X_0, X_1, X_2$ happen to have the form:
\begin{align*}
  X_0 & :              & \U                                                     \\
  X_1 & :              & X_0 \rightarrow X_0 \rightarrow \U                     \\
  X_2 & : \Pi a b c d. & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \\
      &                & \rightarrow X_1 (c)(d) \rightarrow \U                  \\
  \ldots
\end{align*}

>>>>

\usepackage{amsmath, amssymb, mathrsfs, wasysym, url, bookmark, tikz, tikz-cd, mathpazo, enumitem, xargs, multirow, float, tabularx, environ, booktabs, setspace, makecell, colortbl, pdflscape, flushend, minted, art.cls/colorpal, art.cls/ct, art.cls/sset, art.cls/lim, art.cls/joinargs}

@@

\subsection*{Fibered vs indexed presentation of semi-simplicial and semi-cubical sets}
A family of sets can commonly be represented in two ways: as a family properly speaking, indexed over a given set $S$ or as a set $T$ together with a map specifying for each element of the set $T$ on which index in $S$ it depends. In the first case, we call it an \emph{indexed} presentation. In the latter case, the set associated to a given index is the fiber, i.e. inverse image, of the index under the map, so we call it a \emph{fibered} presentation. The two presentations are equivalent and the equivalence can be phrased concisely in the language of homotopy type theory~\cite{hottbook} as a fibered/indexed correspondence\footnote{At a discussion on the Zulip channel for the HoTT community, alternative names fibration/family correspondence and unbundled/bundled correspondence were proposed, among others. In our experience, the opposition fibered/indexed is spontaneously understood by many listeners. In particular, it echoes to the Grothendieck construction of fibered categories from indexed categories. The most elementary instance of the correspondence, with $\Type$ instead of $\U$, that is $(\Sigma S: \Type. (S \rightarrow T)) \simeq (T \rightarrow \Type)$, is sometimes called ``Grothendieck construction for dummies'' (its proof requires univalence, see~\cite{hottbook}).}
\begin{equation*}
  \mbox{(fibered)}\qquad(\Sigma S: \U. (S \rightarrow T)) ~\simeq~ (T \rightarrow \U)\qquad \mbox{(indexed)}
\end{equation*}
where $\U$ represents sets in homotopy type theory as the subset of types with a unique proof of equality (within some universe).

@@

In category-theoretic terms, the indexing set of the family $X_n$ in the indexed presentation is called a \emph{matching object}. The indexed presentation of presheaves over a direct category can then be identified with a subclass of ordinary presheaves known as Reedy fibrant where Reedy fibrancy precisely expresses that the set $X_n$ in the fibered presentation is a $\Sigma$-type over an instance up to isomorphism of the matching object at level $n$~\citep{shulman14,kraus17}.

@@

@@

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj_{\DeltaPlus}       & := \Nat                                                                             \\
      \Hom_{\DeltaPlus}(p, n) & := \{l \in \{0, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}              \\
      g \circ f               & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                        \\
        0 :: (g' \circ f)  & \text{if $g = (0 :: g')$}                                                 \\
        a :: (g' \circ f') & \text{if $g = (\star :: g')$, $f = (a :: f')$, where $a = 0$ or $\kstar$} \\
      \end{cases}                \\
      \id                     & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom_{\DeltaPlus}(n, n)$}
    \end{align*}
  \end{figure}
\end{definition}

@@

\begin{definition}[$\DeltaPlus^n$]
  The standard augmented $n$-semi-simplex $\DeltaPlus^n$ is defined as the Yoneda embedding of $n \in \Obj(\DeltaPlus)$:
  \begin{align*}
    \DeltaPlus^n    & : \Set_{\DeltaPlus}        \\
    \DeltaPlus^n(p) & := \Hom(p, n)              \\
    \DeltaPlus^n(f) & := \lambda g .\, g \circ f
  \end{align*}
\end{definition}

@@

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj_{\Cube}       & := \Nat                                                                          \\
      \Hom_{\Cube}(p, n) & := \{l \in \{L, R, \kstar\}^n \mid \text{number of $\kstar$ in $l = p$}\}        \\
      g \circ f          & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                               \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a = L$ or $R$}                           \\
        a :: (g' \circ f') & \text{if $g = (\kstar :: g')$, $f = (a :: f')$, where $a = L$, $R$, or $\kstar$} \\
      \end{cases} \\
      \id                & := [\kstar, \ldots, \kstar] \text{ $n$ times}
    \end{align*}
  \end{figure}

@@

\begin{example}[$\Cube^1$]
  $\Cube^1$ consists of two points, given by $\Hom(0, 1)$, and a line, given by $\Hom(1, 1)$.
  \begin{equation*}
    \begin{tikzcd}
      L \arrow[r, "\kstar", dash] & R
    \end{tikzcd}
  \end{equation*}
\end{example}

@@

\subsection*{\texorpdfstring{$\nu$}{ν}-sets}
Let us call $\nu$-sets, the generalization of augmented semi-simplicial sets and semi-cubical sets. To obtain this, we extend $\DeltaPlus$ and $\Cube$ in a straightforward manner into a category which we call $\hexagon$.

\begin{definition}[$\hexagon$]
  The definition of $\nu$-semi-shape category is shown below. Note that, if $g \circ f$ is well-defined, then the length of $f$ is less than that of $g$. It can be shown that composition is associative and that $\id$ is neutral.

  \begin{figure}[H]
    \centering
    \begin{align*}
      \Obj_{\hexagon}       & := \Nat                                                                           \\
      \Hom_{\hexagon}(p, n) & := \{l \in (\nu \cup \{\kstar\})^n \mid \text{number of $\kstar$ in $l = p$}\}    \\
      g \circ f             & :=
      \begin{cases}
        f                  & \text{if $g = []$}                                                                 \\
        a :: (g' \circ f)  & \text{if $g = (a :: g')$}, \text{where $a \in \nu$}                                \\
        a :: (g' \circ f') & \text{if $g = (\kstar :: g')$, $f = (a :: f')$, where $a \in \nu$ or $a = \kstar$} \\
      \end{cases}   \\
      \id                   & := [\kstar, \ldots, \kstar] \text{ $n$ times for $\id \in \Hom_{\hexagon}(n, n)$}
    \end{align*}
  \end{figure}
\end{definition}

@@

A $\nu$-set is thus a contravariant functor $\phi$ from the $\nu$-semi-shape category to $\Set$ and we call $n$-$\nu$-semi-shape an element of $\phi(n)$. As in the augmented semi-simplicial and semi-cubical cases, the standard $(n + 1)$-$\nu$-semi-shape is obtained by connecting together $\nu$ copies of the standard $n$-$\nu$-semi-shape with an extra copy stretched in the new dimension. We clarify in the next sections how this process of construction is similar to the parametricity translation developed for functional programming~\citep{reynolds83} and more generally for type theory~\citep{bernardy10, bernardy11, atkey14, bernardy15}.

\section{Type theory}
Martin-Löf's Type theory~\citep{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Agda~(\citeyear{agda23}), Coq~(\citeyear{coq22}) or Lean~(\citeyear{lean15}). In type theory, propositions are types and proofs are programs. A particularity of type theory is also that types and programs, hence propositions and proofs also, are considered modulo an equational theory called definitional equality.

Type theory is a flexible formalism supporting different models. Some models are based on topological spaces, where equality is interpreted as path, and substitutivity of equality as transport~\citep{kapulkin21}. These models support the univalence principle stating that equality of types mimics equivalence of types, leading to the development of Homotopy Type Theory~\citep{hottbook}. In type theory, types are organized in a hierarchy of universes written $\Type_m$ for $m$ a natural number. Main types in type theory are the types of dependent pairs, written $\Sigma a : A.\,B(a)$, the types of dependent functions, written $\Pi a:A.\,B(a)$, for $A$ a type and $B(a)$ a type dependent on the inhabitant $a$ of $A$, and the type of propositional equality, written $t = u$. We assume our type theory to also include a distinguished singleton type, written $\unittype$, and with inhabitant $\unitpoint$, the type of boolean values, called $\textsf{bool}$, and the type of natural numbers. We also write $\hd$ and $\tl$ the projections of dependent pairs, and $\refl$ for reflexivity. Logical propositions being types themselves, we use $\Pi$ to represent universal quantification and $\Sigma$ to represent existential quantification.

A type-theoretic notion of sets can be recovered in each universe as $\U[m]$, denoting the subtype of $\Type_m$ for which paths are degenerated, what can be expressed by the property of Uniqueness of Identity Proofs (UIP). Technically, this is expressed as a structure equipping a domain $\Dom$ with the property $\UIP$:
\begin{align*}
  \Dom & : \Type_m                                  \\
  \UIP & : \Pi x y: \Dom.\, \Pi p q: x = y.\, p = q
\end{align*}

@@
